Depression, Anxiety, and Objective Hardship: Examining Prenatal Mental Health During the Covid-19 Pandemic in Canada

### 2024-04-08 outputs/paper.qmd
finalize the paper

---
title: "Depression, Anxiety, and Objective Hardship: Examining Prenatal Mental Health During the Covid-19 Pandemic in Canada"
author: 
  - Boxuan Yi
format: 
  pdf:
    documentclass: article
    geometry: margin = 1in
    abstract: "The outbreak of the Covid-19 pandemic was a substantial stressor, especially for pregnant individuals. Since depression and anxiety beyond moderate levels are considered clinically significant, this paper uses Bayesian model and logistic regression to evaluate the probability of experiencing at least moderate levels of prenatal depression or anxiety during the pandemic. The analysis indicates that younger age, low levels of education and household income, and a high perception of danger to both oneself and the unborn are associated with higher levels of depression and anxiety."
date: "2 April 2024"
date-format: "D MMMM YYYY"
number-sections: true
thanks: "Code and data in this analysis is available at: https://github.com/Elaineyi1/Prenatal_Mental_Health"
bibliography: references.bib
toc: true
---
\newpage
# Introduction {#sec-introduction}
While pregnancy and childbirth are often moments of joy, they can also bring about distress as mothers navigate responsibilities and physical changes. In the 12 months before childbirth, 12.5% of women in Canada reported that most days were very stressful, and 7.5% reported depressive or anxious symptoms [@canadaPregnancyWomenaposs]. These figures are likely to increase during the COVID-19 pandemic due to heightened uncertainty and panic. Maternal depression and anxiety during pregnancy are associated with adverse neurodevelopmental outcomes for the offspring, such as hyperactivity disorder, autism, poor cognitive or emotional development [@MPMHPbeta], which emphasizes the importance of prenatal mental health for both mothers and children. In light of this, I utilize data from the Open Science Framework to explore factors contributing to maternal depression and anxiety during the pandemic, along with associated birth outcomes. This analysis could also provide insight into mothers’ mental health during other objective hardships or the general effects of prenatal distress.

Considering that moderate depression and anxiety can affect daily functioning, this paper distinguishes between levels of distress and use logistic regression to estimate the likelihood of pregnant individuals experiencing at least moderate levels of depression and anxiety in Canada during the pandemic. Predictors include age, education, household income, language preference, and perceived threat to oneself and the unborn baby. The estimand being explored is whether pregnant individuals, during the pandemic, would experience at least moderate levels of depression or anxiety, as self-reported using the Edinburgh Postnatal Depression Scale and the Patient-Reported Outcomes Measurement Information System Scales in the survey.

The results indicate that young pregnant individuals with lower education and household income, as well as a high perception of danger to both themselves and their unborn child, are more likely to experience at least moderate depression and anxiety. Furthermore, participants who completed the survey in French demonstrate a lower probability of having mental distress compared to those who respond in English. Although the dataset used in this paper provides valuable information on pregnant participants’ demographic information, basic birth outcomes, and mental health data, it does not include any information on newborns’ health. Consequently, this paper mainly focuses on the maternal well-being and its effects on mothers.

This paper is constructed as follows: The data utilized for analysis is discussed in @sec-data. Following that, two models, one for depression and one for anxiety, are introduced in @sec-model. @sec-results then contains the results derived from the models. Lastly, I will discuss the findings in a broader context and address limitations of this paper as well as the future explorations in @sec-discussion. This paper uses the programming language R [@r]. The analysis, the model and the visualizations use the following packages: `rstanarm` [@rrstanarm], `modelsummary` [@rmodelsummary], `ggplot2` [@rggplot2], `dplyr` [@rdplyr], `readr` [@rreadr], `janitor` [@rjanitor], `arrow` [@rarrow], `here` [@rhere], `broom` [@rbroom], and `knitr` [@rknitr].

@@

# Data {#sec-data}
## Data Overview and Measurement
The prenatal mental health dataset sourced from the Open Science Framework was designed to investigate the relationship between exposure to objective hardship caused by the Covid -19 pandemic and psychological distress among pregnant individuals [@osfPregnancyDuring]. The population is pregnant individuals aged 17 or older, with gestation periods not exceeding 35 weeks, residing in Canada. Data collection spanned from May 2020 to December 2021, containing demographic information, mental health data and basic birth outcomes [@nihPrenatalMental]. For this paper, I will only use data collected with delivery month from May to December in 2020 to analyze the prenatal mental health during the initial phase of the outbreak, a period highly marked by panic and adversity.

Participants were recruited through pregnancy organizations, care providers, social media, and paid ads on Facebook and Instagram, with the chance to win a 500-dollar gift card. Some of the advertisements were specifically targeted to geographic regions and sociodemographic groups with lower representation to reduce underrepresentation [@nihPrenatalMental]. Surveys were available in either English or French. Prenatal data, including age, household income in Canadian dollars for the year 2019, and education level were collected as part of the national Pregnancy during the COVID-19 Pandemic (PdP) project using online questionnaires through REDCap. Birth data were also acquired using parents reports [@nihPrenatalMental].

Depression symptoms were self-assessed using the Edinburgh Postnatal Depression Scale (EPDS), comprising 10 questions and each scored from 0 to 3, with possible scores ranging from 0 to 30 [@EDPSstanford]. Anxiety symptoms were self-evaluated using the Patient-Reported Outcomes Measurement Information System (PROMIS), consisting of 7 questions scored from 1 to 5, with possible total scores ranging from 7 to 35 [@psychiatry]. Higher scores on both scales indicate more severe depression or anxiety symptoms. To standardize the scale in this analysis, EPDS scores were divided by 30, and PROMIS scores were adjusted by subtracting 7 from the original scores and then dividing by 35, making the levels of depression and anxiety between 0 and 1.

@@

clean_data |>
  ggplot(aes(x = month, y = PROMIS)) +
  geom_boxplot() + ylim(0,1)
```
\newpage
Firstly, [@fig-month] provides an overview of the levels of depression and anxiety with delivery months spanning from May to December. Overall, a decreasing trend is observed in both scales with the onset of the pandemic. The level of anxiety tends to be slightly higher than that of depression, with the majority experiencing depression levels below 0.5. However, depression exhibits more outliers towards higher values, implying that some individuals experienced significantly severe depression symptoms. Mothers scoring above 0.4 in depression and 0.464 in anxiety are experiencing at least moderate distress. The distributions in [@fig-month] indicate a considerable number of pregnant individuals may be experiencing distress, given that the means fall within the range of 0.3 to 0.5.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-income_age
#| fig-cap: The Distribution of Age Groups and Household Income for Pregnant Participants
# Pastel1
clean_data |>
  ggplot(mapping = aes(x = household_income, fill = age_group)) +
  geom_bar(position = "dodge2") + 
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  labs(x = "Household Income in 2019", y = "Number of Observations", fill = "Age Group") 
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-education_age
#| fig-cap: The Distribution of Age Groups and Level of Education for Pregnant Participants
clean_data |>
  ggplot(mapping = aes(x = maternal_education, fill = age_group)) +
  geom_bar(position = "dodge2") + 
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = "Individual's Education Level", y = "Number of Observations", fill = "Age Group")
```
In Canada, the mean age of mothers at the time of delivery is approximately 31 [@statcanMeanMother]. The dataset contains maternal ages ranging from 18.5 to 49, and because 35 and above is categorized as advanced maternal age [@nihMaternalRisk], I have grouped the ages as follows: Under 27, 27 to 34, 35 to 39, and 40 and above. The distribution of pregnant individuals’ ages and household incomes is illustrated in [@fig-income_age], and the relationship between age and education is depicted in [@fig-education_age]. Both graphs show that a significant proportion of participants fall within the 27 to 34 age group, followed by those aged 35 to 39. Surprisingly, the number of participants under 27 is very low, almost comparable to those aged 40 and above.

A significant proportion of respondents reported incomes ranging from 70,000 to 149,999, as indicated in [@fig-income_age]. There is a notable presence of households with incomes exceeding 200,000, while very few households reported incomes below 40,000. For reference, the median after-tax income of Canadian households was 73,000 in 2020, and 11.1% of Canadians were in the low-income group [@statcanIncomeCanada]. [@fig-education_age] demonstrates that the majority of pregnant individuals in the survey hold undergraduate degrees, with some having completed college/trade school or earned Master’s degrees. Very few respondents reported having less than a high school diploma. According to Statistics Canada, in 2021, 26.3% of women+ aged 15 and above hold a high school diploma, 21.2% had college, CEGEP, or other non-university certificate or diploma. 58.4% hold a postsecondary certificate, while 5.9% completed a Master’s Degree [@statcanHighestLevel].

\newpage
Regarding age and language, [@fig-age_language_dep_anx] presents the proportions of participants experiencing less than moderate levels of depression/anxiety and those experiencing at least moderate depression/anxiety, categorized by age and language. Darker colours represent a more severe level of depression or anxiety. From [@fig-age_language_dep_anx] (a) and (c), younger participants under 27 are more prone to feeling depressed and anxious, with the difference in depression being more noticeable. [@fig-age_language_dep_anx] (b) and (d) reveal that participants who completed the survey in French have lower levels of distress. Given that French speakers are concentrated in Quebec, which was one of the provinces most affected by the pandemic in 2020 [@COVID19Tracker], I consider the language, potentially indicating the region of residence, as a factor influencing the probability of significant distress among pregnant individuals during the Covid-19 pandemic.

```{r}
#| label: fig-age_language_dep_anx
#| fig-cap: Levels of Depression and Anxiety Among Pregnant Individuals Grouped By Age and Language Used to Complete the Survey
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| fig-subcap: ["Age and Depression", "Language and Depression", "Age and Anxiety", "Language and Anxiety"]
clean_data |> 
  ggplot(aes(x=age_group,fill = factor(depression))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "lightcyan3", "1" = "cadetblue"), 
                    labels = c("None or \nMild Depression", "Moderate Depression \nor Higher"))+
  coord_flip()+
  labs(fill = "Depression", x = "Age Group", y = "Proportion") +
  guides(fill = guide_legend(position = "bottom"))

@@

[@fig-income_education_dep_anx] displays the proportions of participants with less than moderate levels of depression/anxiety and those with at least moderate depression/anxiety, categorized by household income in 2019 and education. There is a correlation between higher education and higher household income with lower levels of both depression and anxiety. Conversely, individuals with lower education and income levels are more likely to experience at least moderate depression or anxiety. Comparing [@fig-income_education_dep_anx] (a) and (c), and [@fig-income_education_dep_anx] (b) and (d), we observe that a larger proportion of people experience anxiety compared to depression. More details can be found in @sec-appendix-average.
```{r}
#| label: fig-income_education_dep_anx
#| fig-cap: Levels of Depression and Anxiety Among Pregnant Individuals Grouped By Household Income and Education
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| fig-subcap: ["Income and Depression", "Education and Depression", "Income and Anxiety", "Education and Anxiety"]
clean_data |> 
  ggplot(aes(x=household_income,fill = factor(depression))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "lightcyan3", "1" = "cadetblue"), 
                    labels = c("None or \nMild Depression", "Moderate Depression \nor Higher"))+
  coord_flip()+
  labs(fill = "Depression", x = "Household Income in 2019", y = "Proportion") + 
  guides(fill = guide_legend(position = "bottom"))

@@

clean_data |>
  ggplot(aes(x = maternal_education, fill = factor(anxiety))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "#f4bfb6", "1" = "#f09286"), 
                    labels = c("None or \nMild Anxiety", "Moderate Anxiety \nor Higher")) +
  coord_flip() +
  labs(fill = "Anxiety", x = "Level of Education", y = "Proportion")+
  guides(fill = guide_legend(position = "bottom"))
```
\newpage
Three questions specific to fears of COVID-19 were asked in the survey. Answers were on a numeric scale from 0 (not at all) to 100 (very much so), with 50 representing ‘somewhat’ [@nihPrenatalMental].

@@

Given the similarity between questions 1 and 2, which respectively focus on self-perceived danger and danger perceived towards the unborn, I will analyze the responses to these two questions, but the relationship between question 2 and 3 concerning the baby will be presented in @sec-appendix-questions. Scores are divided by 100, resulting in a range from 0 to 1. The relationship between the perceived level of danger, both towards self and the unborn, and prenatal mental health is visualized in [@fig-danger].

@@

Therefore, the predictors I choose to estimate the prenatal mental health are maternal age groups, household income, education, the perceived danger to oneself, and the perceived danger to the unborn baby.

@@

For the anxiety model, I also expect a positive relationship between experiencing anxiety and age under 27, low levels of education and income, a preference of speaking English, and high perception of danger to both self and unborn baby. This is based on @sec-data as well.

# Results {#sec-results}
## Depression
The coefficients from the logistic regression model, presented in [@tbl-depression_model], show the impact of various predictors on whether a pregnant individual would experience at least moderate depression. Age groups 27 to 34, 35 to 39, and 40 and above all exhibit negative coefficients, suggesting a lower likelihood of depression in individuals aged 27 and above. The general trend in income is that a higher household income positively correlates with a lower probability of depression, 200,000 and above, shows a slightly higher coefficient than the second higher income range of 175,000 to 199,999. Similarly, higher levels of education are associated with lower depression levels, with the highest level, the Doctoral Degree holders, having slightly larger coefficients than those with the second highest degree, the Master’s degrees. Individuals who use French have a negative coefficient, implying a smaller likelihood of depression. The perceived threats to oneself and the unborn baby exhibit large positive coefficients. The threat to life has a larger coefficient than the threat to the baby, meaning that feeling the threat to life can lead to more severe depression.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-depression_model
#| tbl-cap: Estimating Prenatal Depression given Age, Household Income, Education Level and Choice of Language
modelsummary(
  list(
    "Depression" = depression_model))
```
A participant with Master’s Degree, a household income of 175,000 to 199,999, aged 27 to 34, who speaks French, and reports a low perceived threat is least likely to feel depressed. Conversely, a participant with less than a high school diploma, a household income between 20,000 to 39,999, aged under 27, who speaks English, and reports a high perception of danger is the most likely to experience depression.

To provide intuitive insights into the model, I visualize the coefficients from [@tbl-depression_model] in [@fig-depression_interval] along with their 95% confident intervals. The age variables, the language, and threats variables have small confidence intervals. The intercept and the education variables display relatively large confidence intervals, suggesting lower precision in these estimates.

@@

## Anxiety
The coefficients from the logistic regression model for anxiety, presented in [@tbl-anxiety_model], indicate the influence of predictor variables on whether a pregnant individual would experience at least moderate anxiety. Age groups 27 to 34, 35 to 39, and 40 and above have negative coefficients from [@tbl-anxiety_model], suggesting a smaller likelihood of anxiety in individuals aged 27 and above, particularly in the 35 to 39 age group. Higher household income is associated with a lower probability of anxiety, with the income range of 175,000 to 199,999 showing the lowest coefficient. While higher levels of education correspond to lower anxiety, the differences are less evident compared to the depression model. To be more specific, the highest degree Doctoral Degree has coefficient of -0.695, which falls between the coefficients of College/trade school and Undergraduate degree, with coefficients of -0.642 and -0.705 respectively. Pregnant individuals who speak French have a negative coefficient, which implies a lower likelihood of anxiety as well. The perceived threat to oneself and the unborn baby have large positive coefficients, and they are larger than the coefficients in the depression model. The two values are similar, 1.505 and 1.555, respectively. The perception of threat to the baby leads to more severe anxiety symptoms.

A participant with Master’s Degree, a household income of 175,000 to 199,999, aged 35 to 39, who speaks French, and reports a low perceived threat is least likely to experience anxiety. Conversely, a participant with less than a high school diploma, a household income below 20,000, aged under 27, who speaks English, and reports a high perception of danger is the most likely to feel anxious.

@@

>>>>

---
title: "Depression, Anxiety, and Objective Hardship: Examining Prenatal Mental Health During the Covid-19 Pandemic in Canada"
author: 
  - Boxuan Yi
format: 
  pdf:
    documentclass: article
    geometry: margin = 1in
    abstract: "The outbreak of the Covid-19 pandemic was a substantial stressor, especially for pregnant individuals. Since depression and anxiety beyond moderate levels are considered clinically significant, this paper uses the Bayesian model and logistic regression to evaluate the probability of experiencing at least moderate levels of prenatal depression or anxiety during the pandemic. The analysis indicates that younger age, low levels of education and household income, and a high perception of danger to both oneself and the unborn are associated with higher levels of depression and anxiety."
date: "8 April 2024"
date-format: "D MMMM YYYY"
number-sections: true
thanks: "Code and data in this analysis is available at: https://github.com/Elaineyi1/Prenatal_Mental_Health"
bibliography: references.bib
toc: true
---
\newpage
# Introduction {#sec-introduction}
While pregnancy and childbirth are often moments of joy, they can also bring about distress as mothers navigate responsibilities and physical changes. In the 12 months before childbirth, 12.5% of women in Canada reported that most days were very stressful, and 7.5% reported depressive or anxious symptoms [@canadaPregnancyWomenaposs]. These figures were likely to increase during the COVID-19 pandemic due to heightened uncertainty and panic. Maternal depression and anxiety during pregnancy are associated with adverse neurodevelopmental outcomes for the offspring, such as hyperactivity disorder, autism, poor cognitive or emotional development [@MPMHPbeta], which emphasizes the importance of prenatal mental health for both mothers and children. In light of this, I utilize data from the Open Science Framework to explore factors contributing to maternal depression and anxiety during the pandemic, along with associated birth outcomes. This analysis could also provide insight into mothers’ mental health during other objective hardships or the general effects of prenatal distress.

Considering that moderate depression and anxiety can affect daily functioning, this paper distinguishes between levels of distress and uses logistic regression to estimate the likelihood of pregnant individuals experiencing at least moderate levels of depression and anxiety in Canada during the pandemic. Predictors include age, education, household income, language preference, and perceived threat to oneself and the unborn baby. The estimand being explored is whether pregnant individuals, during the pandemic, would experience at least moderate levels of depression or anxiety, as self-reported using the Edinburgh Postnatal Depression Scale and the Patient-Reported Outcomes Measurement Information System Scales in the survey.

The results indicate that young pregnant individuals with lower education and household income, as well as a high perception of danger to both themselves and their unborn child, are more likely to experience at least moderate depression and anxiety. Furthermore, participants who completed the survey in French demonstrate a lower probability of having mental distress compared to those who responded in English. Although the dataset used in this paper provides valuable information on pregnant participants’ demographic information, basic birth outcomes, and mental health data, it does not include any information on newborns’ health. Consequently, this paper mainly focuses on maternal well-being and its effects on mothers.

This paper is constructed as follows: The data utilized for analysis is included in @sec-data. Following that, two models, one for depression and one for anxiety, are introduced in @sec-model. @sec-results then talks about the results derived from the models. Lastly, I will discuss the findings in a broader context and address limitations of this paper as well as the future explorations in @sec-discussion. This paper uses the programming language R [@r]. The analysis, the model and the visualizations use the following packages: `rstanarm` [@rrstanarm], `modelsummary` [@rmodelsummary], `ggplot2` [@rggplot2], `dplyr` [@rdplyr], `readr` [@rreadr], `janitor` [@rjanitor], `arrow` [@rarrow], `here` [@rhere], `broom` [@rbroom], and `knitr` [@rknitr].

@@

# Data {#sec-data}
## Data Overview and Measurement
The prenatal mental health dataset sourced from the Open Science Framework was designed to investigate the relationship between exposure to objective hardship caused by the Covid-19 pandemic and psychological distress among pregnant individuals [@osfPregnancyDuring]. The population is pregnant individuals aged 17 or older, with gestation periods not exceeding 35 weeks, residing in Canada. Data collection spanned from May 2020 to December 2021, containing individual-level demographic information, mental health data and basic birth outcomes [@nihPrenatalMental]. For this paper, I will only use data collected with delivery month from May to December in 2020 to analyze the prenatal mental health during the initial phase of the outbreak, a period highly marked by panic and adversity. 
Participants were recruited through pregnancy organizations, care providers, social media, and paid ads on Facebook and Instagram, with the chance to win a 500-dollar gift card. Some of the advertisements were specifically targeted to geographic regions and sociodemographic groups with lower representation to reduce underrepresentation [@nihPrenatalMental]. Surveys were available in either English or French. Prenatal data, including age, household income in Canadian dollars for the year 2019, and education level were collected as part of the national Pregnancy during the COVID-19 Pandemic (PdP) project using online questionnaires through REDCap. Birth data were also acquired using parents reports [@nihPrenatalMental].

Depression symptoms were self-assessed using the Edinburgh Postnatal Depression Scale (EPDS), comprising ten questions and each scored from 0 to 3, with possible scores ranging from 0 to 30 [@EDPSstanford]. Anxiety symptoms were self-evaluated using the Patient-Reported Outcomes Measurement Information System (PROMIS), consisting of 7 questions scored from 1 to 5, with possible total scores ranging from 7 to 35 [@psychiatry]. Higher scores on both scales indicate more severe depression or anxiety symptoms. To standardize the scale in this analysis, EPDS scores were divided by 30, and PROMIS scores were adjusted by subtracting 7 from the original scores and then dividing by 35, making the levels of depression and anxiety between 0 and 1.

@@

clean_data |>
  ggplot(aes(x = month, y = PROMIS)) +
  geom_boxplot() + ylim(0,1)
```
\newpage
Firstly, [@fig-month] provides an overview of the levels of depression and anxiety with delivery months spanning from May to December. Overall, a decreasing trend is observed in both scales with the onset of the pandemic. The level of anxiety tends to be slightly higher than that of depression, with the majority experiencing depression levels below 0.5. However, depression exhibits more outliers towards higher values, implying that some individuals experienced significantly severe depression symptoms. Mothers scoring above 0.4 in depression and 0.464 in anxiety are experiencing at least moderate distress. The distributions in [@fig-month] indicate a considerable number of pregnant individuals may be experiencing distress, given that the means fall within the range of 0.3 to 0.5.
```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-income_age
#| fig-cap: The Distribution of Age Groups and Household Income for Pregnant Participants
# Pastel1
clean_data |>
  ggplot(mapping = aes(x = household_income, fill = age_group)) +
  geom_bar(position = "dodge2") + 
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  labs(x = "Household Income in 2019", y = "Number of Observations", fill = "Age Group") 
```
\newpage
```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-education_age
#| fig-cap: The Distribution of Age Groups and Level of Education for Pregnant Participants
clean_data |>
  ggplot(mapping = aes(x = maternal_education, fill = age_group)) +
  geom_bar(position = "dodge2") + 
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = "Individual's Education Level", y = "Number of Observations", fill = "Age Group")
```

In Canada, the mean age of mothers at delivery is approximately 31 [@statcanMeanMother]. The dataset contains maternal ages ranging from 18.5 to 49, and because 35 and above is categorized as advanced maternal age [@nihMaternalRisk], I have grouped the ages as follows: Under 27, 27 to 34, 35 to 39, and 40 and above. The distribution of pregnant individuals’ ages and household incomes is illustrated in [@fig-income_age], and the relationship between age and education is depicted in [@fig-education_age]. Both graphs show that a significant proportion of participants fall within the 27 to 34 age group, followed by those aged 35 to 39. Surprisingly, the number of participants under 27 is very low, almost comparable to those aged 40 and above.

A significant proportion of respondents reported incomes ranging from 70,000 to 149,999, as indicated in [@fig-income_age]. There is a notable presence of households with incomes exceeding 200,000, while very few reported incomes below 40,000. For reference, the median after-tax income of Canadian households was 73,000 in 2020, and 11.1% of Canadians were in the low-income group [@statcanIncomeCanada]. [@fig-education_age] demonstrates that the majority of pregnant individuals in the survey hold undergraduate degrees, with some having completed college/trade school or earned Master’s degrees. Very few respondents reported having less than a high school diploma. According to Statistics Canada, in 2021, 26.3% of women+ aged 15 and above held a high school diploma, 21.2% had a college, CEGEP, or other non-university certificate or diploma. 58.4% held a postsecondary certificate, while 5.9% completed a Master’s Degree [@statcanHighestLevel].
\newpage
Regarding age and language, [@fig-age_language_dep_anx] presents the proportions of participants experiencing less than moderate levels of depression/anxiety and those experiencing at least moderate depression/anxiety, categorized by age and language. Darker colours represent a more severe level of depression or anxiety. From [@fig-age_language_dep_anx] (a) and (c), younger participants under 27 are more prone to feeling depressed and anxious, with the difference in depression being more noticeable. [@fig-age_language_dep_anx] (b) and (d) reveal that participants who completed the survey in French have lower levels of distress. Given that French speakers are concentrated in Quebec, which was one of the provinces most affected by the pandemic in 2020 [@COVID19Tracker], I consider the language, potentially indicating the region of residence, as a factor influencing the probability of significant distress among pregnant individuals during the Covid-19 pandemic.
```{r}
#| label: fig-age_language_dep_anx
#| fig-cap: Levels of Depression and Anxiety Among Pregnant Individuals Grouped By Age and Language Used to Complete the Survey
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| fig-subcap: ["Age and Depression", "Language and Depression", "Age and Anxiety", "Language and Anxiety"]
clean_data |> 
  ggplot(aes(x=age_group,fill = factor(depression))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "lightcyan3", "1" = "cadetblue"), 
                    labels = c("None or \nMild Depression", "Moderate Depression \nor Higher"))+
  coord_flip()+
  labs(fill = "Depression", x = "Age Group", y = "Proportion") +
  guides(fill = guide_legend(position = "bottom"))

@@

```{r}
#| label: fig-income_education_dep_anx
#| fig-cap: Levels of Depression and Anxiety Among Pregnant Individuals Grouped By Household Income and Education
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| fig-subcap: ["Income and Depression", "Education and Depression", "Income and Anxiety", "Education and Anxiety"]
clean_data |> 
  ggplot(aes(x=household_income,fill = factor(depression))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "lightcyan3", "1" = "cadetblue"), 
                    labels = c("None or \nMild Depression", "Moderate Depression \nor Higher"))+
  coord_flip()+
  labs(fill = "Depression", x = "Household Income in 2019", y = "Proportion") + 
  guides(fill = guide_legend(position = "bottom"))

@@

clean_data |>
  ggplot(aes(x = maternal_education, fill = factor(anxiety))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("0" = "#f4bfb6", "1" = "#f09286"), 
                    labels = c("None or \nMild Anxiety", "Moderate Anxiety \nor Higher")) +
  coord_flip() +
  labs(fill = "Anxiety", x = "Level of Education", y = "Proportion")+
  guides(fill = guide_legend(position = "bottom"))
```
[@fig-income_education_dep_anx] displays the proportions of participants with less than moderate levels of depression/anxiety and those with at least moderate depression/anxiety, categorized by household income in 2019 and education. There is a correlation between higher education and higher household income with lower levels of both depression and anxiety. Conversely, individuals with lower education and income levels are more likely to experience at least moderate depression or anxiety. Comparing [@fig-income_education_dep_anx] (a) and (c), and [@fig-income_education_dep_anx] (b) and (d), we observe that a larger proportion of people experience anxiety compared to depression. More details can be found in @sec-appendix-average.

\newpage
Three questions specific to fears of COVID-19 were asked in the survey. Answers were on a numeric scale from 0 (not at all) to 100 (very much so), with 50 representing ‘somewhat’ [@nihPrenatalMental].

@@

Given the similarity between questions 1 and 2, which respectively focus on self-perceived danger and danger perceived towards the unborn, I will analyze the responses to these two questions, while the relationship between questions 2 and 3 concerning the baby will be presented in @sec-appendix-questions. Scores are divided by 100, resulting in a range from 0 to 1. The relationship between the perceived level of danger, both towards self and the unborn, and prenatal mental health is visualized in [@fig-danger].

@@

Therefore, the predictors I choose to estimate prenatal mental health are maternal age groups, household income, education, the perceived danger to oneself, and the perceived danger to the unborn baby.

@@

For the anxiety model, I also expect a positive relationship between experiencing anxiety and age under 27, low levels of education and income, a preference of speaking English, and a high perception of danger to both self and unborn baby. This is based on @sec-data as well.

# Results {#sec-results}
## Depression
The coefficients from the logistic regression model, presented in [@tbl-depression_model], show the impact of various predictors on whether a pregnant individual would experience at least moderate depression. Age groups 27 to 34, 35 to 39, and 40 and above all exhibit negative coefficients, suggesting a lower likelihood of depression in individuals aged 27 and above. The general trend in income is that a higher household income positively correlates with a lower probability of depression, 200,000 and above, shows a slightly higher coefficient than the second higher income range of 175,000 to 199,999. Similarly, higher levels of education are associated with lower depression levels, with the highest level, the Doctoral Degree holders, having slightly larger coefficients than those with the second highest degree, the Master’s degree. Individuals who use French have a negative coefficient, implying a smaller likelihood of depression. The perceived threats to oneself and the unborn baby exhibit large positive coefficients. The threat to life has a larger coefficient than the threat to the baby, meaning that feeling the threat to life can lead to more severe depression.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-depression_model
#| tbl-cap: Estimating Prenatal Depression given Age, Household Income, Education Level and Choice of Language
modelsummary(
  list(
    "Depression" = depression_model))
```

A participant with a Master’s Degree, a household income of 175,000 to 199,999, aged 27 to 34, who speaks French, and reports a low perceived threat is least likely to feel depressed. Conversely, a participant with less than a high school diploma, a household income between 20,000 to 39,999, aged under 27, who speaks English, and reports a high perception of danger is the most likely to experience depression.

To provide intuitive insights into the model, I visualize the coefficients from [@tbl-depression_model] in [@fig-depression_interval] along with their 95% confidence intervals. The age, the language, and threats variables have small confidence intervals. The intercept and the education variables display relatively large confidence intervals, suggesting lower precision in these estimates.

@@

## Anxiety
The coefficients from the logistic regression model for anxiety, presented in [@tbl-anxiety_model], indicate the influence of predictor variables on whether a pregnant individual would experience at least moderate anxiety. Age groups 27 to 34, 35 to 39, and 40 and above have negative coefficients from [@tbl-anxiety_model], suggesting a smaller likelihood of anxiety in individuals aged 27 and above, particularly in the 35 to 39 age group. Higher household income is associated with a lower probability of anxiety, with the income range of 175,000 to 199,999 showing the lowest coefficient. While higher levels of education correspond to lower anxiety, the differences are less evident compared to the depression model. To be more specific, the highest degree Doctoral Degree has a coefficient of -0.695, which falls between the coefficients of College/trade school and Undergraduate degree, with coefficients of -0.642 and -0.705 respectively. Pregnant individuals who speak French have a negative coefficient, which implies a lower likelihood of anxiety as well. The perceived threats to oneself and the unborn baby have large positive coefficients, and they are larger than the coefficients in the depression model. The two values are similar, 1.505 and 1.555, respectively. The perception of threat to the baby leads to more severe anxiety symptoms.

@@

A participant with a Master’s Degree, a household income of 175,000 to 199,999, aged 35 to 39, who speaks French, and reports a low perceived threat is least likely to experience anxiety. Conversely, a participant with less than a high school diploma, a household income below 20,000, aged under 27, who speaks English, and who reports a high perception of danger is the most likely to feel anxious.
